---
title: "Comprehensive Competing Risks Analysis of aGVHD"
author: "ml4time2event team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Competing Risks Analysis of aGVHD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This vignette illustrates the competing-risks workflow on the aGVHD dataset: build a task, fit registered engines, obtain tidy predictions/metrics, and visualize with standard verbs.

# 1. Packages and data

First, we load the necessary libraries and the dataset. We will use a sample of the data to speed up the analysis.

```{r load-data}
library(ml4time2event)
library(dplyr)
library(ggplot2)
library(recipes)
library(data.table) # Add the `data.table` library to ensure `fread` is available

# Load the data
full_data <- fread(file.path("..", "data", "agvhd24_individual_hla.csv"))

# Take a sample for this vignette to run faster
set.seed(123)
sample_indices <- sample(nrow(full_data), size = nrow(full_data))
data_raw <- full_data[sample_indices, ]

## Choose event of interest (1 = aGVHD 2-4, 2 = death)
event_of_interest <- 1L

# We'll create binary `status` and `cause` columns after splitting so we can
# ensure the training split contains the chosen event. If not, we'll pick an
# available non-zero event automatically.
```

Create a simple hold-out split for validation.

```{r split}
set.seed(123)
# remove rows of data with missing in target variables
data_raw <- data_raw |>
  filter(!is.na(agvhd_2_4_time), !is.na(agvhd_2_4_event))

# convert comorbidity_score to numeric
data_raw <- data_raw |>
  mutate(comorbidity_score = as.numeric(comorbidity_score))

# Convert multi-class status into binary status and a separate cause column, matching the comprehensive vignette
# The key is: status becomes binary (0=censored, 1=any event), cause captures the original event type
# Important: rows with NA in agvhd_2_4_event should have NA in both status and cause
data <- data_raw |>
  filter(!is.na(agvhd_2_4_event)) |>
  mutate(
    cause = if_else(agvhd_2_4_event == 0L, NA_integer_, as.integer(agvhd_2_4_event)),
    status = if_else(agvhd_2_4_event == 0L, 0L, 1L)
  )

# Now split the transformed data
train_rows <- sample.int(nrow(data), size = floor(0.7 * nrow(data)))
data_train <- data[train_rows, ]
data_test  <- data[-train_rows, ]

# Impute missing values in predictors using recipes
# Exclude outcome columns from imputation
exclude_cols <- c("agvhd_2_4_time", "agvhd_2_4_event", "status", "cause")
predictor_cols <- setdiff(colnames(data_train), exclude_cols)

# Create recipe for imputation
impute_recipe <- recipe(~ ., data = data_train) |>
  update_role(all_of(exclude_cols), new_role = "outcome") |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors())

# Fit recipe on training data
impute_recipe_prep <- prep(impute_recipe, training = data_train)

# Apply to both training and test data
data_train <- bake(impute_recipe_prep, new_data = data_train)
data_test <- bake(impute_recipe_prep, new_data = data_test)

sprintf("Training rows: %s  |  Validation rows: %s",
        nrow(data_train), nrow(data_test))
table(data_train$status, data_train$cause)
table(data_train$status, data_train$agvhd_2_4_event)

```

# 2. Build a competing-risk task

```{r task-definition}
# Create the competing risks task
t2e_task <- ml4t2e_task_cr(
  data = data_train,
  time = "agvhd_2_4_time",
  status = "status",
  cause = "cause",
  features = setdiff(colnames(data_train), c("agvhd_2_4_time", "agvhd_2_4_event", "status", "cause"))
)

print(t2e_task)
```

# 3. Fit registered engines (with ensembling)

The registry exposes a consistent set of engine identifiers; you can list them via `ml4t2e_list_models("competing_risk")`.

```{r fit-models, message=FALSE, warning=FALSE}
# List available models and fit them
ml4t2e_list_models("competing_risk")

# Fit models and create an ensemble (match comprehensive vignette)
cr_fit <- ml4t2e_fit(
  task = t2e_task,
  models = c("cox", "fine_gray", "cr_xgboost"),
  ensemble = "auto",
  controls = list(times = seq(0, max(data_train$agvhd_2_4_time, na.rm = TRUE), length.out = 60))
)

cr_fit
```

Training metrics are again a single function call away.

```{r train-metrics}
ml4t2e_evaluate(
  cr_fit,
  metrics = c("c_index", "ibs"),
  include = c("ensemble", "cox", "fine_gray", "cr_xgboost")
)
```

# 4. Score a hold-out set

```{r score}
# Create task for test data
t2e_task_test <- ml4t2e_task_cr(
  data = data_test,
  time = "agvhd_2_4_time",
  status = "status",
  cause = "cause",
  features = setdiff(colnames(data_test), c("agvhd_2_4_time", "agvhd_2_4_event", "status", "cause"))
)

# Get predictions on test set
cr_preds_val <- predict(
  cr_fit,
  newdata = data_test,
  times = seq(0, max(data_test$agvhd_2_4_time, na.rm = TRUE), length.out = 60),
  type = "cif",
  include = c("ensemble","cox", "fine_gray", "cr_xgboost")
)

ml4t2e_evaluate(
  cr_preds_val,
  task = t2e_task_test,
  metrics = c("c_index", "ibs")
)
```

# 5. Visualise cumulative incidence curves

`autoplot()` aggregates model-specific CIFs across individuals and separates them by cause.

```{r autoplot-fit, fig.width=7, fig.height=4}
autoplot(cr_fit, include = c("ensemble","cox", "fine_gray", "cr_xgboost"))
```

```{r autoplot-preds, fig.width=7, fig.height=4}
autoplot(cr_preds_val, include = "ensemble")
```

# Variable Importance
