---
title: "Comprehensive Survival Analysis with ml4time2event"
author: "ml4time2event Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Comprehensive Survival Analysis Pipeline with ml4time2event

This vignette demonstrates an end-to-end survival analysis workflow using the
**ml4time2event** package. We rely on the prepared NCCTG lung cancer dataset
distributed with the package, build an ensemble of survival models, evaluate
performance, and visualise predictions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Setup and Load Libraries

```{r load-libraries}
library(ml4time2event)
library(survival)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

# 2. Load Prepared Dataset

The dataset `lung_survival_prepared` ships with the package and contains the
imputed and feature-engineered variables used throughout this vignette.

```{r load-data}
lung_data <- ml4time2event::get_lung_survival_data()

glimpse(lung_data)

summary(lung_data[c("time", "status", "age", "ph.karno", "wt.loss")])
```

# 3. Explore Baseline Characteristics

We review the event distribution and follow-up horizon to understand the sample.

```{r explore-data}
status_counts <- dplyr::count(lung_data, status)
status_counts

event_rate <- mean(lung_data$status == 1)
time_range <- range(lung_data$time)

data.frame(
  metric = c("Event rate", "Minimum follow-up (days)", "Maximum follow-up (days)"),
  value = c(round(event_rate, 3), round(time_range[1], 1), round(time_range[2], 1))
)
```

# 4. Variable Profiling with Package Tools

`VariableProfile()` provides quick summaries for candidate predictors.

```{r variable-profile}
candidate_expvars <- c(
  "age", "sex", "ph.ecog", "ph.karno", "pat.karno",
  "meal.cal", "wt.loss", "age_group", "performance_good"
)
var_profile <- VariableProfile(lung_data, candidate_expvars)
var_profile$Summary
```

# 5. Data Splitting

We create a reproducible 70/30 train-test split using `t2edata_split()`.

```{r split-data}
#set.seed(123)
split_data <- t2edata_split(lung_data, prop = 0.7)
train_data <- split_data[["Train"]]
test_data <- split_data[["Test"]]

data.frame(
  dataset = c("Training", "Test"),
  observations = c(nrow(train_data), nrow(test_data))
)
```

# 6. Survival Model Training

`RunSurvModels()` trains multiple survival learners and prepares the ensemble
object used for downstream prediction.

```{r train-models}
timevar <- "time"
eventvar <- "status"
expvars <- c("age", "sex", "ph.ecog", "ph.karno", "meal.cal", "wt.loss")

data.frame(
  setting = c("Time variable", "Event variable", "Predictors"),
  value = c(timevar, eventvar, paste(expvars, collapse = ", "))
)

models <- RunSurvModels(
  datatrain = train_data,
  ExpVars = expvars,
  timevar = timevar,
  eventvar = eventvar,
  models = c("glmnet", "coxph", "rulefit", "xgboost", "gam", "gbm", "ExpSurvReg", "bart", "deepsurv"),
  ntreeRF = 300,
  nvars = 20,
  ntimes = 50  # New parameter: number of time points for prediction grid
)

# Individual model components now include:
# - For Cox model: ntimes, alpha (elastic net), nfolds (cross-validation)
# - For XGBoost: ntimes (prediction grid), verbose (progress reporting)
# - Consistent time grid specification across all models

models
```

# 7. Ensemble Predictions

We generate survival probability curves for the held-out test set and compute an
average ensemble across models. All individual models now use a consistent time
grid from their training phase, which is automatically propagated to predictions.

```{r generate-predictions}
complete_test_data <- test_data[complete.cases(test_data[, expvars]), ]

# Set prediction time grid to match observed event times for correct concordance calculation

# Define actual_times and actual_events for prediction grid
actual_times <- complete_test_data[[timevar]]
actual_events <- complete_test_data[[eventvar]]

# Use sorted unique observed event times for prediction grid
prediction_times <- sort(unique(actual_times))

ensemble_predictions <- PredictSurvModels(
  models = models,
  newdata = complete_test_data,
  new_times = prediction_times,
  ensemble_method = "average"
)

individual_predictions <- ensemble_predictions$ModelPredictions
ensemble_probs <- ensemble_predictions$NewProbs
common_times <- ensemble_predictions$NewTimes
models_used <- ensemble_predictions$models_used

data.frame(
  metric = c(
    "Test subjects (complete cases)",
    "Prediction time points",
    "Models in ensemble"
  ),
  value = c(
    nrow(complete_test_data),
    length(common_times),
    paste(models_used, collapse = ", ")
  )
)
```

# 7. Organise Model Predictions

```{r prepare-predictions}
pretty_model_name <- function(x) {
  mapping <- c(
    RF_Model = "Random Forest",
    RF_Model2 = "Random Forest (Top Vars)",
    glmnet_Model = "GLMNet",
    CPH_Model = "Cox PH",
    bart_Model = "BART",
    deepsurv_Model = "DeepSurv",
    gam_Model = "GAM",
    gbm_Model = "GBM",
    survregexp_Model = "ExpSurvReg",
    survregweib_Model = "WeibSurvReg",
    xgboost_Model = "XGBoost",
    RuleFit_Model = "RuleFit"
  )
  if (x %in% names(mapping)) mapping[[x]] else tools::toTitleCase(gsub("_", " ", gsub("_Model$", "", x)))
}

model_prediction_objects <- lapply(individual_predictions, function(mat) {
  list(Times = common_times, Probs = mat)
})
names(model_prediction_objects) <- vapply(names(individual_predictions), pretty_model_name, character(1))
all_predictions <- c(
  model_prediction_objects,
  list(Ensemble = list(Times = common_times, Probs = ensemble_probs))
)
names(all_predictions)
```

# 8. Model Evaluation

We evaluate every model (including the ensemble) using time-dependent concordance,
Brier score, and mean expected time lost. The updated API provides consistent 
metric calculation across survival and competing risks models.

```{r evaluate-models}
actual_times <- complete_test_data[[timevar]]
actual_events <- complete_test_data[[eventvar]]
actual_events_binary <- as.integer(actual_events == 1)

median_time <- median(actual_times[actual_events_binary == 1], na.rm = TRUE)
max_time <- as.numeric(quantile(actual_times, 0.9, na.rm = TRUE))

etl_results <- list()

performance_summary <- do.call(rbind, lapply(names(all_predictions), function(model_name) {
  pred_obj <- all_predictions[[model_name]]

 
  # Calculate ETL values ONCE and reuse
  etl_values <- tryCatch({
    etl_list <- CalculateExpectedTimeLost(
      PredictedCurves = list(list(NewProbs = pred_obj$Probs)),
      modeltypes = c("SURV"),
      times = pred_obj$Times,
      UL = max_time,
      quiet = TRUE  # Suppress verbose output
    )
    etl_list[[1]]
  }, error = function(e) {
    warning("Error calculating ETL for model ", model_name, ": ", e$message)
    rep(NA_real_, ncol(pred_obj$Probs))
  })

  # Store for later use
  etl_results[[model_name]] <<- etl_values

  # Concordance using package function
  concordance <- tryCatch({
    # Use integratedC for survival models, integratedConcordanceCR for CR models
    if (model_name %in% c("RuleFit", "Random Forest", "Random Forest (Top Vars)", "GLMNet", "Cox PH", "BART", "DeepSurv", "GAM", "GBM", "ExpSurvReg", "XGBoost", "Ensemble")) {
        integratedC(
          predsurv = pred_obj$Probs,
          pred_times = pred_obj$Times,
          obstimes = actual_times,
          obsevents = actual_events_binary
        )
    } else {
        integratedConcordanceCR(
          SurvObj = survival::Surv(actual_times, actual_events),
          Predictions = pred_obj$Probs,
          eval_times = pred_obj$Times,
          cause = 1
        )
    }
  }, error = function(e) NA_real_)

  # BrierScore
  brier <- tryCatch({
    brier_obj <- BrierScore(
      predsurv = pred_obj$Probs,
      pred_times = pred_obj$Times,
      obstimes = actual_times,
      obsevents = actual_events_binary,
      eval_times = median_time
    )
    if (!is.null(brier_obj$AppErr)) brier_obj$AppErr$model[1] else NA_real_
  }, error = function(e) NA_real_)

  data.frame(
    Model = model_name,
    Concordance = concordance,
    Brier_Score = brier,
    Mean_ETL = mean(etl_values, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}))

performance_summary

```

# 9. Expected Time Lost Analysis

Expected time lost (ETL) summarises the area between the survival curve and full
follow-up horizon. The table below highlights the 90th percentile follow-up time
and the ensemble ETL distribution.

```{r expected-time-lost}
ensemble_etl <- etl_results[["Ensemble"]]

data.frame(
  statistic = c(
    "90th percentile follow-up (days)",
    "Mean ensemble ETL (days)",
    "Ensemble ETL range (days)"
  ),
  value = c(
    round(max_time, 1),
    round(mean(ensemble_etl, na.rm = TRUE), 2),
    paste(
      round(range(ensemble_etl, na.rm = TRUE), 2),
      collapse = " – "
    )
  )
)
```

# 10. Visualisation

We visualise the ensemble survival curves for a few patients and inspect the ETL
distribution.

```{r visualization, fig.width=10, fig.height=8}

ordered_models <- performance_summary[order(-performance_summary$Concordance), ]
ordered_models <- ordered_models[!is.na(ordered_models$Concordance), ]

# Plot all models with ensemble highlighted in black
plot_predictions <- all_predictions

p_surv <- plot_survival_curves(
  predictions = plot_predictions,
  patients_to_plot = seq_len(min(3, ncol(ensemble_probs))),
  highlight_ensemble = TRUE,
  title = "Predicted Survival Curves - All Models (Ensemble in Black)"
)
print(p_surv)

ensemble_etl_clean <- ensemble_etl[!is.na(ensemble_etl)]
if (length(ensemble_etl_clean) > 0) {
  etl_df <- data.frame(
    Subject = seq_along(ensemble_etl_clean),
    ETL = ensemble_etl_clean
  )

  p_etl <- ggplot(etl_df, aes(x = ETL)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "black") +
    labs(
      title = "Ensemble Expected Time Lost Distribution",
      x = "Expected Time Lost (days)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))

  print(p_etl)
}
```

# 11. Model Persistence

`SaveEnsemble()` and `LoadEnsemble()` make it easy to persist fitted ensembles.
We store objects in temporary files for illustration.

```{r save-models}
ensemble_file <- tempfile("lung_cancer_ensemble_", fileext = ".rds")
results_file <- tempfile("lung_cancer_ensemble_results_", fileext = ".rds")

SaveEnsemble(models, file = ensemble_file)

ensemble_results <- list(
  predictions = ensemble_predictions,
  individual_predictions = individual_predictions,
  ensemble_probs = ensemble_probs,
  times = common_times,
  evaluation = performance_summary,
  etl_results = etl_results
)
saveRDS(ensemble_results, results_file)

loaded_ensemble <- LoadEnsemble(file = ensemble_file)
loaded_results <- readRDS(results_file)

data.frame(
  object = c("Loaded ensemble class", "Stored individual model predictions"),
  value = c(
    paste(class(loaded_ensemble), collapse = ", "),
    length(loaded_results$individual_predictions)
  )
)
```

# 12. Key Takeaways

## Updated API Features

- **Standardized time grid management**: The `ntimes` parameter controls prediction grid resolution across all models (default: 50 time points). Individual models (Cox, XGBoost) now manage time grids consistently.

- **Enhanced Cox model**: `SurvModel_Cox()` now supports `ntimes`, `alpha` (elastic net regularization), and `nfolds` (cross-validation folds) for more flexible regularization.

- **XGBoost improvements**: `SurvModel_xgboost()` now includes `ntimes` (prediction grid) and `verbose` (progress reporting) parameters for better control and monitoring.

- **Unified metric API**: `BrierScore()` and `BrierScoreCR()` now use the consistent `eval_times` parameter (instead of `time` for scalar, `eval_times` for vector support) for seamless survival and competing risks model evaluation.

- **Automatic time attribute propagation**: Predictions now include a `Times` attribute for automatic time grid tracking through the analysis pipeline.

## Workflow Summary

- Prepared dataset `lung_survival_prepared` (available through `get_lung_survival_data()`) streamlines preprocessing for the vignette.
- `RunSurvModels()` and `PredictSurvModels()` coordinate multiple survival learners and ensemble predictions with consistent time grid management.
- Evaluation utilities cover concordance, Brier score, and expected time lost across individual models and the ensemble with unified API.
- Visualisation and persistence helpers make it straightforward to operationalise survival models.

## Migration Notes

- If you have existing code using `time` parameter in metrics, the package maintains backward compatibility by accepting both `time` (scalar) and `eval_times` (vector).
- Individual model predictions are now consistently structured with `Times` attribute for easy access to prediction time points.
