---
title: "Comprehensive Survival Analysis"
author: "ml4time2event team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

The refactored `ml4time2event` package centres on three tidy verbs:

1. create a task with `ml4t2e_task_surv()` or `ml4t2e_task_cr()`
2. fit engines through `ml4t2e_fit()`
3. work with tidy tibbles returned from `predict()`, `ml4t2e_evaluate()`, and `autoplot()`.

This vignette walks through an end-to-end survival workflow that showcases the new API. We will:

* explore the built-in lung cancer dataset
* train a collection of survival engines (with optional ensembling)
* score a hold-out set and compare metrics
* visualise survival curves and permutation-based feature importance
* wrap everything into a reusable pipeline (recipes + resampling)

# 1. Packages and data

```{r libraries}
library(ml4time2event)
library(dplyr)
library(ggplot2)
library(survival)
```

The NCCTG lung-cancer dataset ships with the package. We convert the event indicator to the 0/1 format expected by the new task constructors.

```{r data}
lung_df <- get_lung_survival_data()
lung_df$id <- seq_len(nrow(lung_df))

if (is.data.frame(lung_df$status)) {
  lung_df$status <- lung_df$status[[1]]
}

status_max <- max(lung_df$status, na.rm = TRUE)
if (status_max > 1L) {
  # Fall back to 0/1 coding when the source data uses 1/2 indicators
  lung_df$status <- ifelse(lung_df$status == status_max, 1L, 0L)
}

lung_df$status <- as.integer(lung_df$status)

glimpse(lung_df)
```

For the remainder of the analysis we keep a simple training / validation split.

```{r split}
# set.seed(123)
train_rows <- sample.int(nrow(lung_df), size = floor(0.7 * nrow(lung_df)))
lung_train <- lung_df[train_rows, ]
lung_test <- lung_df[-train_rows, ]

sprintf(
  "Training rows: %s  |  Validation rows: %s",
  nrow(lung_train), nrow(lung_test)
)
```

# 2. Build a survival task

Tasks capture the outcome columns, the feature set, and lightweight metadata.

```{r task}
feature_cols <- c(
  "age", "sex", "ph.ecog", "ph.karno", "pat.karno",
  "meal.cal", "wt.loss"
)

surv_task <- ml4t2e_task_surv(
  data = lung_train,
  time = "time",
  event = "status",
  features = feature_cols,
  id = "id",
  time_units = "days"
)
```

# 3. Fit multiple engines (with optional ensemble)

`ml4t2e_fit()` pulls constructor definitions from the internal registry. The `controls` list lets us share hyperparameters (e.g. the common prediction grid).

```{r fit}
surv_fit <- ml4t2e_fit(
  task = surv_task,
  models = c("cox", "random_forest"),
  ensemble = "auto",
  controls = list(times = seq(0, 1000, length.out = 75)),
  keep_data = TRUE
)

surv_fit
```

Evaluation on the training data is one function call away.

```{r train-metrics}
ml4t2e_evaluate(
  surv_fit,
  metrics = c("c_index", "ibs"),
  include = c("ensemble", "cox", "random_forest")
)
```

# 4. Score a validation set



```{r validate}
surv_task_val <- ml4t2e_task_surv(
  data = lung_test,
  time = "time",
  event = "status",
  features = feature_cols,
  id = "id",
  time_units = "days"
)

surv_preds_val <- predict(
  surv_fit,
  newdata = lung_test,
  times = seq(0, 1000, length.out = 75),
  type = "survival",
  include = c("ensemble", "cox", "random_forest")
)

ml4t2e_evaluate(
  surv_preds_val,
  task = surv_task_val,
  metrics = c("c_index", "ibs"),
  include = c("ensemble", "cox", "random_forest")
)
```

# 5. Visualise survival curves

`autoplot()` works on both fitted objects and tidy prediction tibbles.

```{r autoplot-fit-v2, fig.width=7, fig.height=4}
autoplot(surv_fit, include = c("ensemble", "cox", "random_forest"))
```

```{r autoplot-preds, fig.width=7, fig.height=4}
autoplot(surv_preds_val, include = "ensemble")
```

# 6. Permutation-based feature importance

Permutation importance is built into the core API for survival fits.

```{r explain, fig.width=7, fig.height=4}
surv_importance <- ml4t2e_explain(
  surv_fit,
  newdata = lung_test,
  top_n = 6
)

surv_importance

autoplot(surv_importance)
```

# 7. Pipelines with recipes and resampling

For workflows that require preprocessing and resampling, wrap everything in a `T2EPipeline`. Recipes are optional, but when supplied they are baked automatically during prediction/evaluation.

```{r pipeline}
library(recipes)
library(rsample)

surv_pipeline <- ml4t2e_pipeline(
  outcome = list(type = "survival", time = "time", event = "status"),
  models = c("cox", "random_forest", "gbm", "xgboost"),
  ensemble = "auto",
  recipe = recipe(~., data = lung_train) |>
    step_impute_median(all_numeric_predictors()) |>
    step_dummy(all_nominal_predictors()),
  resampling = vfold_cv(lung_train, v = 3)
)

surv_pipeline$fit(lung_train)
surv_pipeline$summary()
```

Pipelines expose the same prediction and evaluation helpers:

```{r pipeline-eval}
surv_pipeline$evaluate(
  newdata = lung_test,
  metrics = c("c_index", "ibs"),
  include = "ensemble"
)
```

# 8. Persist results

Fits are regular R objects; use `ml4t2e_save()` / `ml4t2e_load()` for the primary API or base R `saveRDS()` for ad-hoc serialization.

```{r persistence}
tmp_fit <- tempfile(fileext = ".rds")
ml4t2e_save(surv_fit, tmp_fit)
reloaded_fit <- ml4t2e_load(tmp_fit)

tmp_pipe <- tempfile(fileext = ".rds")
saveRDS(surv_pipeline, tmp_pipe)
reloaded_pipeline <- readRDS(tmp_pipe)

reloaded_fit
reloaded_pipeline$summary()
```

# Conclusion

The new object-oriented surface keeps common analyses compact:

1. declare a task,
2. fit one or more registry-backed engines (optionally with ensembling),
3. reuse the same tidy outputs across prediction, evaluation, visualisation, and explainability.

The pipeline helpers layer on recipes/resampling without sacrificing the simplicity of the core verbs.
