---
title: "Comprehensive Competing-Risks Analysis with ml4time2event"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Competing Risks Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

The competing-risks path mirrors the survival UX introduced in the refactor: build a task, fit registered engines, obtain tidy predictions/metrics, and visualise or explain with standard verbs. This vignette illustrates the workflow on the classic bone marrow transplant (BMT) dataset.

# 1. Packages and data

```{r libraries}
library(ml4time2event)
library(dplyr)
library(ggplot2)
```

`get_bmt_competing_risks_data()` returns the prepared dataset that was used in the legacy API. We only need to split the multi-class status into a binary event flag plus a separate cause column.

```{r data}
bmt_raw <- get_bmt_competing_risks_data()

bmt_df <- bmt_raw |>
  mutate(
    cause = if_else(status == 0L, NA_integer_, status),
    status = if_else(status == 0L, 0L, 1L)
  )

glimpse(bmt_df)
table(bmt_df$status, useNA = "ifany")
```

Create a simple hold-out split for validation.

```{r split}
set.seed(123)
train_rows <- sample.int(nrow(bmt_df), size = floor(0.7 * nrow(bmt_df)))
bmt_train <- bmt_df[train_rows, ]
bmt_test <- bmt_df[-train_rows, ]

sprintf(
  "Training rows: %s  |  Validation rows: %s",
  nrow(bmt_train), nrow(bmt_test)
)
```

# 2. Build a competing-risk task

```{r task}
cr_task <- ml4t2e_task_cr(
  data = bmt_train,
  time = "ftime",
  status = "status",
  cause = "cause",
  features = c("sex", "d", "phase", "age", "source"),
  time_units = "years"
)
```

The task stores the time range, the cause mapping, and the feature set that downstream engines rely upon.

# 3. Fit registered engines (with ensembling)

The registry exposes a consistent set of engine identifiers; you can list them via `ml4t2e_list_models("competing_risk")`.

```{r fit}
ml4t2e_list_models("competing_risk")

cr_fit <- ml4t2e_fit(
  task = cr_task,
  models = c("cox", "cr_fine_gray", "cr_random_forest", "cr_shallownn", "cr_ttah"),
  ensemble = "auto",
  controls = list(times = seq(0, 120, length.out = 60)),
  keep_data = TRUE
)

cr_fit
```

Training metrics are again a single function call away.

```{r train-metrics}
ml4t2e_evaluate(
  cr_fit,
  metrics = c("c_index", "ibs"),
  include = c("ensemble", "cox", "cr_fine_gray", "cr_random_forest", "cr_shallownn", "cr_ttah")
)
```

# 4. Score a hold-out set

```{r validate}
cr_task_val <- ml4t2e_task_cr(
  data = bmt_test,
  time = "ftime",
  status = "status",
  cause = "cause",
  features = c("sex", "d", "phase", "age", "source"),
  time_units = "years"
)

cr_preds_val <- predict(
  cr_fit,
  newdata = bmt_test,
  times = seq(0, 120, length.out = 60),
  type = "cif",
  include = c("ensemble", "cox", "fine_gray", "cr_random_forest", "cr_shallownn", "cr_ttah")
)

ml4t2e_evaluate(
  cr_preds_val,
  task = cr_task_val,
  metrics = c("c_index", "ibs")
)
```

# 5. Visualise cumulative incidence curves

`autoplot()` aggregates model-specific CIFs across individuals and separates them by cause.

```{r autoplot-fit, fig.width=7, fig.height=4}
autoplot(cr_fit, include = c("ensemble", "cox", "cr_random_forest", "cr_shallownn", "cr_ttah"))
```

```{r autoplot-preds, fig.width=7, fig.height=4}
autoplot(cr_preds_val, include = "ensemble")
```

# 6. Pipelines with preprocessing and resampling

Pipelines provide the same interface for competing risks. The example below fits two engines with a simple recipe (median imputation + dummy encoding) and three-fold cross-validation.

```{r pipeline}
library(recipes)
library(rsample)

cr_pipeline <- ml4t2e_pipeline(
  outcome = list(
    type = "competing_risk",
    time = "ftime",
    status = "status",
    cause = "cause"
  ),
  models = c("cox", "cr_fine_gray", "cr_random_forest", "cr_shallownn", "cr_ttah"),
  ensemble = "auto",
  recipe = recipe(status ~ ., data = bmt_train) |>
    step_impute_median(all_numeric_predictors()) |>
    step_dummy(all_nominal_predictors()),
  resampling = vfold_cv(bmt_train, v = 3)
)

cr_pipeline$fit(bmt_train)
cr_pipeline$summary()
```

Evaluating on the validation set reuses the same helper:

```{r pipeline-eval}
cr_pipeline$evaluate(
  newdata = bmt_test,
  metrics = c("c_index", "ibs"),
  include = "ensemble"
)
```

Predictions are tidy tibbles, ready for `dplyr` or `ggplot2`.

```{r pipeline-predict}
cr_pipeline$predict(
  newdata = bmt_test[1:4, ],
  times = seq(0, 60, by = 6),
  type = "cif",
  include = "ensemble"
)
```

# 7. Persist fitted objects

Use `ml4t2e_save()` / `ml4t2e_load()` for `t2e_fit` objects and base R `saveRDS()` for pipelines.

```{r persistence}
tmp_fit <- tempfile(fileext = ".rds")
ml4t2e_save(cr_fit, tmp_fit)
ml4t2e_load(tmp_fit)

tmp_pipe <- tempfile(fileext = ".rds")
saveRDS(cr_pipeline, tmp_pipe)
readRDS(tmp_pipe)$summary()
```

# Summary

The competing-risks UX now matches the survival path: define a task, fit registered engines, obtain tidy predictions/metrics, and layer in preprocessing/resampling through pipelines when needed. Ensembles are controlled through the same `ensemble` argument, and the outputs stay consistent across survival and competing-risk workflows.
