---
title: "Comprehensive Competing Risks Analysis with ml4time2event"
author: "ml4time2event Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Competing Risks Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Comprehensive Competing Risks Analysis with ml4time2event

This vignette walks through a complete competing-risks workflow with
**ml4time2event**. Using the prepared bone marrow transplant dataset, we train
an ensemble of competing-risks models, evaluate their performance, and produce
visualisations tailored to cumulative incidence analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Setup and Load Libraries

```{r load-libraries}
library(ml4time2event)
library(survival)
library(dplyr)
library(ggplot2)

if (!requireNamespace("cmprsk", quietly = TRUE)) {
  stop("Install the 'cmprsk' package to run this vignette.")
}
library(cmprsk)
```

# 2. Load Prepared Dataset

The dataset `bmt_competing_risks_prepared` is bundled with the package and
accessible via `get_bmt_competing_risks_data()`, which also provides a fallback
preparation routine during development.

```{r load-data}
bmt_data <- ml4time2event::get_bmt_competing_risks_data()

glimpse(bmt_data)

summary(bmt_data[c("ftime", "status", "age")])
```

# 3. Explore Event Distribution

```{r explore-events}
# Event counts: 0=censored, 1=relapse, 2=treatment-related mortality
table(bmt_data$status)
```

# 4. Variable Profiling with Package Tools

```{r variable-profile}
candidate_expvars <- c("sex", "d", "phase", "age", "source")
var_profile <- VariableProfile(bmt_data, candidate_expvars)
var_profile
```

# 5. Data Splitting

```{r split-data}
set.seed(123)
split_data <- t2edata_split(bmt_data, prop = 0.7)
train_data <- split_data$Train
test_data <- split_data$Test
```

# 6. Competing-Risks Model Training

```{r train-models}
timevar <- "ftime"
eventvar <- "status"
expvars <- c("sex", "d", "phase", "age", "source")
primary_event_code <- 1L

models <- RunCRModels(
  datatrain = train_data,
  ExpVars = expvars,
  timevar = timevar,
  eventvar = eventvar,
  models = c("cox", "FG", "xgboost", "gam", "bart", "rulefit", "survreg", "ttah", "shallownn")
)

models
```

# 7. Predictions

```{r generate-predictions}
predictions <- PredictCRModels(
  models = models,
  newdata = test_data,
  new_times = seq(0, max(test_data$ftime), length.out = 50),
  ensemble_method = "average"
)
```

# 8. Model Evaluation

```{r evaluate-ensemble}
leaderboard <- EvaluateCRModels(
  models = models,
  data = test_data,
  timevar = timevar,
  eventvar = eventvar,
  eval_times = seq(0, max(test_data[[timevar]]), length.out = 50),
  ensemble_method = "average",
  cause = primary_event_code
)

leaderboard$display_name <- format_model_name(leaderboard$model, model_type = "competing_risks")
# Display full leaderboard
leaderboard_display <- leaderboard[, c("display_name", "integrated_c", "integrated_brier", "rank")]
knitr::kable(leaderboard_display, row.names = FALSE,
             caption = "Model Performance Leaderboard (ranked by integrated Brier score)")
```

# 9. Visualisation

```{r visualization, fig.width=10, fig.height=6}
plot_cif_curves(
  predictions = predictions,
  patients_to_plot = 1:3,
  highlight_ensemble = TRUE
)
```

# 10. Model Persistence

```{r save-models}
# Save ensemble
ensemble_file <- tempfile("cr_ensemble_model_", fileext = ".rds")
SaveEnsemble(models, file = ensemble_file)

# Load ensemble
loaded_ensemble <- LoadEnsemble(file = ensemble_file)
print(loaded_ensemble)
```
