---
title: "Comprehensive Survival Analysis with ml4time2event"
author: "ml4time2event Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Comprehensive Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Comprehensive Survival Analysis Pipeline with ml4time2event

This vignette demonstrates an end-to-end survival analysis workflow using the
**ml4time2event** package. We rely on the prepared NCCTG lung cancer dataset
distributed with the package, build an ensemble of survival models, evaluate
performance, and visualise predictions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Setup and Load Libraries

```{r load-libraries}
library(ml4time2event)
library(survival)
library(dplyr)
library(ggplot2)
```

# 2. Load Prepared Dataset

The dataset `lung_survival_prepared` ships with the package and contains the
imputed and feature-engineered variables used throughout this vignette.

```{r load-data}
lung_data <- ml4time2event::get_lung_survival_data()

glimpse(lung_data)

summary(lung_data[c("time", "status", "age", "ph.karno", "wt.loss")])
```

# 3. Explore Baseline Characteristics

```{r explore-data}
# Event distribution
table(lung_data$status)

# Summary statistics
summary(lung_data[c("time", "status")])
```

# 4. Variable Profiling with Package Tools

`VariableProfile()` provides quick summaries for candidate predictors.

```{r variable-profile}
candidate_expvars <- c(
  "age", "sex", "ph.ecog", "ph.karno", "pat.karno",
  "meal.cal", "wt.loss", "age_group", "performance_good"
)
var_profile <- VariableProfile(lung_data, candidate_expvars)
var_profile

lung_data$ph.ecog <- as.numeric(lung_data$ph.ecog)
```

# 5. Data Splitting

```{r split-data}
split_data <- t2edata_split(lung_data, prop = 0.7)
train_data <- split_data$Train
test_data <- split_data$Test
```

# 6. Survival Model Training

```{r train-models}
timevar <- "time"
eventvar <- "status"
expvars <- c("age", "sex", "ph.ecog", "ph.karno", "meal.cal", "wt.loss")

models <- RunSurvModels(
  datatrain = train_data,
  ExpVars = expvars,
  timevar = timevar,
  eventvar = eventvar,
  models = c("glmnet", "coxph", "rulefit", "xgboost", "gam", "gbm", "ExpSurvReg", "bart", "shallownn", "ttah"),
  ntimes = 50
)

models
```

# 7. Predictions

```{r generate-predictions}
complete_test_data <- test_data[complete.cases(test_data[, expvars]), ]
prediction_times <- sort(unique(complete_test_data[[timevar]]))

# Generate predictions
predictions <- PredictSurvModels(
  models = models,
  newdata = complete_test_data,
  new_times = prediction_times,
  ensemble_method = "average"
)
```

# 8. Model Evaluation

```{r evaluate-models}
leaderboard <- EvaluateSurvModels(
  models = models,
  data = complete_test_data,
  timevar = timevar,
  eventvar = eventvar,
  eval_times = prediction_times,
  prediction_times = prediction_times,
  ensemble_method = "average"
)

leaderboard$display_name <- format_model_name(leaderboard$model, model_type = "survival")
print(leaderboard[, c("display_name", "integrated_c", "integrated_brier", "rank")])
```

# 9. Visualisation

```{r visualization, fig.width=10, fig.height=6}
plot_survival_curves(
  predictions = predictions,
  patients_to_plot = 1:3,
  highlight_ensemble = TRUE
)
```

# 10. Model Persistence

```{r save-models}
# Save ensemble
ensemble_file <- tempfile("lung_cancer_ensemble_", fileext = ".rds")
SaveEnsemble(models, file = ensemble_file)

# Load ensemble
loaded_ensemble <- LoadEnsemble(file = ensemble_file)
print(loaded_ensemble)
```
