---
title: "SHAP-Based Explainability for Time-to-Event Pipelines"
author: "ml4time2event team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{SHAP-Based Explainability for Time-to-Event Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)
```

# Introduction

The refactored `ml4time2event` exposes tidy predictions for both survival and competing-risk outcomes. `ml4t2e_calculate_shap()` layers expected-time-lost SHAP values on top of those predictions so you can interpret ensemble pipelines with a consistent grammar:

1. build a pipeline via `ml4t2e_pipeline()`
2. compute SHAP values with `ml4t2e_calculate_shap()`
3. visualise them using the helper plots (`ml4t2e_shap_importance()`, `ml4t2e_shap_dependence()`, `ml4t2e_shap_waterfall()`)

Expected time lost integrates the event probability up to a user-defined horizon. For survival models this is $\int_0^T (1 - S(t))\, dt$; for competing risks it is $\int_0^T \text{CIF}(t)\, dt$. The result is a scalar that SHAP can decompose exactly into feature-level contributions.

We rely on `fastshap` for Monte-Carlo SHAP estimation.

```{r libraries}
library(ml4time2event)
library(dplyr)
library(ggplot2)
library(fastshap)
```

# Survival pipeline example

## Fit pipeline

```{r surv-pipeline}
lung <- get_lung_survival_data()

if (is.data.frame(lung$status)) {
  lung$status <- lung$status[[1]]
}

status_max <- max(lung$status, na.rm = TRUE)
if (status_max > 1L) {
  lung$status <- ifelse(lung$status == status_max, 1L, 0L)
}

lung$status <- as.integer(lung$status)

surv_pipe <- ml4t2e_pipeline(
  outcome = list(type = "survival", time = "time", event = "status"),
  models = c("cox", "random_forest"),
  ensemble = "auto"
)

surv_pipe$fit(lung)
surv_pipe$summary()
```

## Compute SHAP values

```{r surv-shap}
set.seed(42)
surv_shap <- ml4t2e_calculate_shap(
  pipeline = surv_pipe,
  data = lung[1:60, ],
  time_horizon = 365,
  nsim = 50
)

surv_shap
```

## Visualise

```{r surv-importance}
ml4t2e_shap_importance(surv_shap, plot_type = "bar", max_features = 8)
```

```{r surv-beeswarm}
ml4t2e_shap_importance(surv_shap, plot_type = "beeswarm", max_features = 8)
```

```{r surv-dependence}
ml4t2e_shap_dependence(surv_shap, feature = "age")
```

```{r surv-waterfall}
ml4t2e_shap_waterfall(surv_shap, obs_id = 5, max_features = 8)
```

# Competing-risk pipeline example

## Fit pipeline

```{r cr-pipeline}
bmt <- get_bmt_competing_risks_data() |>
  mutate(
    cause = if_else(status == 0L, NA_integer_, status),
    status = if_else(status == 0L, 0L, 1L)
  )

cr_pipe <- ml4t2e_pipeline(
  outcome = list(
    type = "competing_risk",
    time = "ftime",
    status = "status",
    cause = "cause"
  ),
  models = c("cox", "fine_gray"),
  ensemble = "auto"
)

cr_pipe$fit(bmt)
cr_pipe$summary()
```

## Compute SHAP values

```{r cr-shap}
set.seed(99)
cr_shap <- ml4t2e_calculate_shap(
  pipeline = cr_pipe,
  data = bmt[1:80, ],
  time_horizon = 100,
  nsim = 40
)

cr_shap
```

## Visualise

```{r cr-importance}
ml4t2e_shap_importance(cr_shap, plot_type = "bar", max_features = 8)
```

```{r cr-dependence}
ml4t2e_shap_dependence(cr_shap, feature = "age")
```

```{r cr-waterfall}
ml4t2e_shap_waterfall(cr_shap, obs_id = 3, max_features = 6)
```

# Tips

* Increase `nsim` for smoother SHAP estimates once you move beyond vignette-sized examples.
* The helper returns the baseline prediction (expected time lost with all features off); this makes it easy to confirm the additivity constraint: prediction = baseline + Σ SHAP.
* For larger pipelines, subset the data before calling `ml4t2e_calculate_shap()` to keep runtimes manageable.

The SHAP helpers respect the tidy outputs of the new API, so they work equally well with survival or competing-risk pipelines.
