---
title: "Testing All Survival Models with a Simple Example"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Testing All Survival Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This vignette demonstrates testing all available survival models in the `ml4time2event` package with a simple example. We will use a small subset of the lung cancer dataset to quickly fit, predict, and evaluate all survival models, ensuring the complete pipeline works correctly.

# 1. Packages and data

```{r libraries}
library(ml4time2event)
library(dplyr)
library(ggplot2)
library(survival)
```

Load the lung cancer dataset and create a small subset for testing:

```{r data}
# Load the full dataset
lung_df <- get_lung_survival_data()

# Create a small subset for quick testing (100 observations)
set.seed(42)
small_lung <- lung_df %>%
  select(time, status, age, sex, ph.ecog, ph.karno, wt.loss)

# Check the data structure
str(small_lung)
summary(small_lung)
```

# 2. Available Survival Models

List all available survival models:

```{r list_models}
# List all survival models
survival_models <- ml4t2e_list_models(outcome = "survival")
survival_models
```

# 3. Create Survival Task

Create a survival analysis task:

```{r task}
# Create survival task
task <- ml4t2e_task_surv(
  data = small_lung,
  time = "time",
  event = "status",
  features = c("age", "sex", "ph.ecog", "ph.karno", "wt.loss")
)

task
```

# Fit All Survival Models

Fit all available survival models (excluding GBM which has issues with small datasets):

```{r fit_models}
models_to_fit <- survival_models$engine
models_to_fit <- setdiff(models_to_fit, "gbm") # Exclude GBM explicitly

# Fit all survival models
fitted_models <- ml4t2e_fit(
  task = task,
  models = models_to_fit,
  seed = 42,
  keep_data = TRUE
)

# Check the fitted models
fitted_models
```

# 5. Make Predictions

Generate predictions for all models:

```{r predictions}
# Make predictions on training data
predictions <- predict(fitted_models)

# Check prediction structure
head(predictions)
summary(predictions)
```

# 6. Evaluate Models

Evaluate the performance of all models:

```{r evaluation}
# Evaluate models
evaluation <- ml4t2e_evaluate(fitted_models)

# Check evaluation results
evaluation
```

# 7. Visualize Results

Create a simple visualization of survival curves for a few models:

```{r visualization}
# Plot survival curves for first few models
plot_data <- predictions %>%
  filter(model %in% c("cox", "random_forest", "xgboost")) %>%
  filter(id == 1) # Focus on one individual

ggplot(plot_data, aes(x = time, y = surv, color = model)) +
  geom_line() +
  labs(
    title = "Survival Curves for Selected Models (Sample Individual)",
    x = "Time",
    y = "Survival Probability",
    color = "Model"
  ) +
  theme_minimal()
```

# Summary

This vignette successfully demonstrated:

- Loading and preparing a small dataset for testing
- Listing all available survival models
- Creating a survival analysis task
- Fitting all survival models simultaneously (excluding GBM due to small dataset limitations)
- Generating predictions
- Evaluating model performance
- Visualizing results

All survival models in the `ml4time2event` package work correctly with this example (GBM requires larger datasets for proper OOB/CV estimation).
<parameter name="filePath">/Users/dakdemir/Library/CloudStorage/OneDrive-NMDP/Year2025/Github/ml4time2event/vignettes/testing_all_survival_models.Rmd