---
title: "BART Implementation - Quick Reference"
subtitle: "Meeting Notes for Dr. Rodney Sparapani"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: false
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

We use the CRAN BART package (v2.9.9) in our `ml4time2event` R package for clinical time-to-event prediction. We implement:
- `BART::surv.bart()` for survival analysis
- `BART::crisk.bart()` for competing risks analysis (both causes in data, predicting CIF for one cause at a time)

# Current Hyperparameters

## Survival BART (`surv.bart`)

```{r}
K = 8          # Time grid points
ntree = 50     # Number of trees
ndpost = 200   # Posterior draws
nskip = 50     # Burn-in
keepevery = 2  # Thinning
```

## Competing Risks BART (`crisk.bart`)

We use `crisk.bart` which handles both causes in the data, but we predict the cumulative incidence function (CIF) for **one cause at a time**.

```{r}
K = 10          # Time grid points
ntree = 100     # Number of trees
ndpost = 1000   # Posterior draws
nskip = 100     # Burn-in
keepevery = 10  # Thinning
numcut = 2      # Cutpoints (for speed)
```

**Note**: Our implementation fits one crisk.bart model per event of interest. The model sees all causes in the data (coded as 0, 1, 2, etc.) but returns CIF predictions for the specified cause only.

---

# Implementation Questions

## 1. Prediction Reshaping
BART returns predictions as a concatenated vector. We reshape using `matrix(..., byrow = TRUE)`. Is this correct?

## 2. Time Ordering
`bart_model$times` are not always sorted. We sort post-hoc with `order()`. Is this the recommended approach?

## 3. Interpolation
We use linear interpolation for custom time points via `survprobMatInterpolator()`. Is this appropriate for BART curves?

## 4. Fitting Failures
`crisk.bart()` occasionally fails. We retry up to 4 times. What causes these failures, and how can we improve stability?

---

# Suggestions for Improvement

## Hyperparameters
- Are our defaults appropriate for clinical datasets (n=100-10,000, p=10-100)?
- Should we adjust based on sample size, censoring rate, or feature dimensionality?
- What are minimum recommended values for stable predictions?

## Implementation
- Is there a better way to extract predictions in matrix format?
- Should we implement joint modeling for multiple competing events?
- Any best practices for clinical prediction tasks?

---

# References

1. Sparapani, R., et al. (2016). Nonparametric survival analysis using BART. *Statistics in Medicine*, 35(16), 2741-2753.
2. Sparapani, R., et al. (2020). Nonparametric competing risks analysis using BART. *Statistical Methods in Medical Research*, 29(1), 57-77.
